Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2022-06-18 response-to-reviews.Rmd

l.313 'information from adjacent mature leaves or apical meristems, including a “memory” of previous conditions in that location, ...'.  This was not in the previous version that I can see, and I don't see the basis for inclusion of this statement. As well, use of the term 'memory' seems awkward if not entirely wrong (even reading between quotes). Please revise.

**Revised as suggested.**

l.579 'implies that high 𝑇𝑎𝑖𝑟 should decrease 𝑔𝑠 of upper canopy leaves more than understory leaves'. The authors here are ignoring what they’ve said previously, that high Tair is associated with high VPD (l. 199-200), and that VPD is conclusively a much stronger driver of gs than Tleaf (Leuning 1995, Grossiord). I think the sentence should be corrected and include VPD.  I'm not yet aware of a good study that has decoupled VPD and Tleaf to prove that Tleaf intrinsically causes reduction in gs. So the authors cannot conclude this.

**We have removed the quoted sentence as it was redundant with that which had preceded it, but had confusingly couched the message of the preceding sentence as if it were a speculation rather than an empirical fact. The preceding sentence, which we have retained, read:**
*“…sun leaves in the upper canopy show a stronger decrease in* *g~s~* *in response to rising* *T~air~* *(driven by increased VPD) than lower-canopy shade leaves (Hernández et al., 2020; Carter et al., 2021)”.*
**That is an empirical result. The fact that the effect of *T~air~* on *g~s~* in that result is driven by VPD rather than somehow directly by *T~air~* itself is indicated by the fact that increasing temperature itself apparently *opens* stomata, rather than closing them. This is also germane to the reviewer’s final comment about decoupling VPD and *T~leaf~*. There are in fact six published studies that have decoupled VPD and *T~leaf~*, and they all suggest temperature per se causes an increase, not a reduction, in stomatal conductance. We have added the following sentence to the end of the paragraph to clarify this matter, citing three of the studies out of six due to word count limit: “Available data also suggest that high *T~air~* can directly drive stomatal opening, if vapor pressure is experimentally adjusted in tandem with *T~air~* to keep VPD constant [@schulzeRoleAirHumidity1974; @hallRegulationWaterLoss1975; @hallStomatalResponseEnvironment1975; @fredeenTemperatureHumidityEffects1999a; @mottStomatalResponsesHumidity2010; @urbanStomatalConductanceIncreases2017a], though it appears that the negative effect of temperature on *g~s~*, mediated by VPD, generally overrides the positive effect of temperature per se in nature, causing *g~s~* to decline when the leaf warms.**

@@

*response required*

l.629 'we currently lack mechanistic understanding of how and why the temperature sensitivity of respiration varies across the forest vertical profile'. Other than that it varies according to the prevailing temperature difference? I think the temperature difference between upper and lower canopy positions, the theme of this review, should be mentioned in regard to this.

**Thanks for suggesting. We included the following sentence:**
**Overall, although *Q~10~* is influenced by the temperature difference between upper and lower canopy positions, we currently lack mechanistic understanding of how and why *Q~10~* varies across the forest vertical profile.**

### 2022-06-18 response-to-reviews.Rmd

l.313 'information from adjacent mature leaves or apical meristems, including a “memory” of previous conditions in that location, ...'.  This was not in the previous version that I can see, and I don't see the basis for inclusion of this statement. As well, use of the term 'memory' seems awkward if not entirely wrong (even reading between quotes). Please revise.

**Revised as suggested.**

l.579 'implies that high 𝑇𝑎𝑖𝑟 should decrease 𝑔𝑠 of upper canopy leaves more than understory leaves'. The authors here are ignoring what they’ve said previously, that high Tair is associated with high VPD (l. 199-200), and that VPD is conclusively a much stronger driver of gs than Tleaf (Leuning 1995, Grossiord). I think the sentence should be corrected and include VPD.  I'm not yet aware of a good study that has decoupled VPD and Tleaf to prove that Tleaf intrinsically causes reduction in gs. So the authors cannot conclude this.

**We have removed the quoted sentence as it was redundant with that which had preceded it, but had confusingly couched the message of the preceding sentence as if it were a speculation rather than an empirical fact. The preceding sentence, which we have retained, read:**
*“…sun leaves in the upper canopy show a stronger decrease in* *g~s~* *in response to rising* *T~air~* *(driven by increased VPD) than lower-canopy shade leaves (Hernández et al., 2020; Carter et al., 2021)”.*
**That is an empirical result. The fact that the effect of *T~air~* on *g~s~* in that result is driven by VPD rather than somehow directly by *T~air~* itself is indicated by the fact that increasing temperature itself apparently *opens* stomata, rather than closing them. This is also germane to the reviewer’s final comment about decoupling VPD and *T~leaf~*. There are in fact six published studies that have decoupled VPD and *T~leaf~*, and they all suggest temperature per se causes an increase, not a reduction, in stomatal conductance. We have added the following sentence to the end of the paragraph to clarify this matter, citing three of the six studies due to limit on word count: “Available data also suggest that high *T~air~* can directly drive stomatal opening, if vapor pressure is experimentally adjusted in tandem with *T~air~* to keep VPD constant [@schulzeRoleAirHumidity1974; @hallRegulationWaterLoss1975; @hallStomatalResponseEnvironment1975; @fredeenTemperatureHumidityEffects1999a; @mottStomatalResponsesHumidity2010; @urbanStomatalConductanceIncreases2017a], though it appears that the negative effect of temperature on *g~s~*, mediated by VPD, generally overrides the positive effect of temperature per se in nature, causing *g~s~* to decline when the leaf warms.**

### 2022-06-18 response-to-reviews.Rmd

l.629 'we currently lack mechanistic understanding of how and why the temperature sensitivity of respiration varies across the forest vertical profile'. Other than that it varies according to the prevailing temperature difference? I think the temperature difference between upper and lower canopy positions, the theme of this review, should be mentioned in regard to this.

**Thanks for suggesting. We included the following sentence:**
**Overall, although the temperature difference between upper and lower canopy positions is important driver of *Q~10~*, we currently lack mechanistic understanding of how and why *Q~10~* varies across the forest vertical profile.**

### 2022-06-18 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

Elevation of *T~leaf~* above *T~air~* varies with micrometeorological conditions. 
During drought (associated with conditions that are hot and dry, with higher-than-average solar radiation), when *g~s~* is limiting, *T~leaf~* - *T~air~* will tend to be higher in the upper canopy compared to the understory (Fig. `r fig_leaf_T`).
This is because sun leaves are exposed to higher irradiance and VPD, and therefore have stronger *g~s~* limitation than shade leaves [Fig. `r fig_leaf_T`e, @leighInfluenceLeafSize2017; @fauset_differences_2018]. 
In contrast, under conditions conducive to stomatal opening, higher wind speeds and *g~b~* enable cooling in the upper canopy, whereas lower wind and *g~b~* in the understory limit evaporative cooling  [Fig. `r fig_leaf_T`b,d, @robertsStomatalBoundaryLayerConductances1990; @martin_boundary_1999; @leighInfluenceLeafSize2017 <!-- @songTreeSurfaceTemperature2020-->]. 
Thus, while understory and within-canopy shade leaves can remain cooler under lower radiation, their environment is less conducive to dissipating excess heat compared to upper-canopy leaves. 
Limitations in heat dissipation in shade leaves, together with physiological propensities, may cause above-optimal *T~leaf~* under slight heat or drought stress or during sunflecks [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020; @schymanskiStomatalControlLeaf2013].
In addition to lower wind speeds, higher RH (lower VPD) in the understory and inner canopy also inhibits evaporative cooling, thus increasing *T~leaf~* - *T~air~* [Fig. `r fig_leaf_T`c, @tibbittsHumidityPlants1979; @songTreeSurfaceTemperature2020;  @perezIncreasingHumidityThreatens2018; @konradLeafTemperatureIts2021a; @dietzVerticalPatternsDuration2007].

@@

Similar to photosynthesis, respiration tends to be higher in upper-canopy sun leaves, but its temperature sensitivity (*Q~10~* or *E~0~*) shows no definite trend along the vertical gradient.
Specifically, the temperature sensitivity of respiration can be constant within vertical profiles and in seedling sun vs. shade leaves<!--[<!--@carterExperimentalWarmingTropical2021-> @weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007 <!--@zaragoza-castellsClimatedependentVariationsLeaf2008; @xuSeasonalVariationTemperature2006->]-->, greater in upper-canopy leaves<!--[@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996]-->, or greater in the lower canopy <!--[@griffinCanopyPositionAffects2002]--> (Table `r table_leaf_metabolism`). 
This variation may be attributable to forest type, leaf traits and age [e.g., greater *Q~10~* in younger leaves, @zhouLeafageEffectsTemperature2015], or acclimation to high temperature that decreases *Q~10~* [@slotGeneralPatternsAcclimation2015; @carterExperimentalWarmingTropical2021]. 
Overall, although *Q~10~* is influenced by the temperature difference between upper and lower canopy positions, we currently lack mechanistic understanding of how and why *Q~10~* varies across the forest vertical profile.

>>>>

Elevation of *T~leaf~* above *T~air~* varies with micrometeorological conditions. 
During drought (associated with conditions that are hot and dry, with higher-than-average solar radiation), when *g~s~* is limiting, *T~leaf~* - *T~air~* will tend to be higher in the upper canopy compared to the understory (Fig. `r fig_leaf_T`).
This is because sun leaves are exposed to higher irradiance and VPD, and therefore have stronger *g~s~* limitation than shade leaves [Fig. `r fig_leaf_T`e, @leighInfluenceLeafSize2017; @fauset_differences_2018]. 
In contrast, under conditions conducive to stomatal opening, higher wind speeds and *g~b~* enable cooling in the upper canopy, whereas lower wind and *g~b~* in the understory limit evaporative cooling  [Fig. `r fig_leaf_T`b,d, @robertsStomatalBoundaryLayerConductances1990; @martin_boundary_1999; @leighInfluenceLeafSize2017 <!-- @songTreeSurfaceTemperature2020-->]. 
Thus, while understory and within-canopy shade leaves can remain cooler under lower radiation, their environment is less conducive to dissipating excess heat compared to upper-canopy leaves, because wind speeds and hence *g~b~* are generally smaller in the lower canopy [@robertsStomatalBoundaryLayerConductances1990; @martin_boundary_1999]. 
Limitations in heat dissipation in shade leaves, together with physiological propensities, may cause above-optimal *T~leaf~* under slight heat or drought stress or during sunflecks [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020; @schymanskiStomatalControlLeaf2013].
In addition to lower wind speeds, higher RH (lower VPD) in the understory and inner canopy also inhibits evaporative cooling, thus increasing *T~leaf~* - *T~air~* [Fig. `r fig_leaf_T`c, @tibbittsHumidityPlants1979; @songTreeSurfaceTemperature2020;  @perezIncreasingHumidityThreatens2018; @konradLeafTemperatureIts2021a; @dietzVerticalPatternsDuration2007].

@@

Similar to photosynthesis, respiration tends to be higher in upper-canopy sun leaves, but its temperature sensitivity (*Q~10~* or *E~0~*) shows no definite trend along the vertical gradient.
Specifically, the temperature sensitivity of respiration can be constant within vertical profiles and in seedling sun vs. shade leaves<!--[<!--@carterExperimentalWarmingTropical2021-> @weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007 <!--@zaragoza-castellsClimatedependentVariationsLeaf2008; @xuSeasonalVariationTemperature2006->]-->, greater in upper-canopy leaves<!--[@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996]-->, or greater in the lower canopy <!--[@griffinCanopyPositionAffects2002]--> (Table `r table_leaf_metabolism`). 
This variation may be attributable to forest type, leaf traits and age [e.g., greater *Q~10~* in younger leaves, @zhouLeafageEffectsTemperature2015], or acclimation to high temperature that decreases *Q~10~* [@slotGeneralPatternsAcclimation2015; @carterExperimentalWarmingTropical2021]. 
Overall, although the temperature difference between upper and lower canopy positions is important driver of *Q~10~*, we currently lack mechanistic understanding of how and why *Q~10~* varies across the forest vertical profile.
